Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-10-31 forecast/paper/supplement-text.tex
DONE with pass over supp

As explained in Section 2.B of the main paper, a (somewhat cynical) hypothesis
for why we see benefits in forecasting and hotspot prediction is that the
indicators are not actually providing useful information but they are instead
acting as a sort of ``implicit regularization,'' leading to shrinkage on the
autoregressive coefficients and therefore to less volatile predictions.  To
investigate this hypothesis, we consider fitting ``noise features'' that in
truth have zero relationship to the response.  Recall (from the main paper) that
at each forecast date, we train a model on 6,426 location-time pairs.  Each
indicator model uses 6 features, corresponding to the 3 autoregressive terms and
the 3 lagged indicator values.  To form noise indicator features, we replace
their values with those from a randomly chosen time-space pair (while keeping
the autoregressive features fixed).  In particular, at each location $\ell$ and
time $t$, for the forecasting task we replace the triplet \smash{$(X_{\ell,t},
X_{\ell,t-7}, X_{\ell,t-14})$} in Eq.\ (3) of the main
paper with the triplet \smash{$(X_{\ell^*,t^*}, X_{\ell^*,t^*-7},
  X_{\ell^*,t^*-14})$}, where $(\ell^*,t^*)$ is a location-time pair sampled with
replacement from the 6,426 total location-time pairs.  Likewise in the hotspot 
prediction task, we replace the triplet \smash{$(X_{\ell,t}^\Delta,
  X_{\ell,t-7}^\Delta, X_{\ell,t-14}^\Delta)$} in Eq.\ (5) of the main paper with
\smash{$(X_{\ell^*,t^*}^\Delta, X_{\ell^*,t^*-7}^\Delta,
  X_{\ell^*,t^*-14}^\Delta)$}.  Figures
\ref{fig:fcast-booted}--\ref{fig:hot-booted} show the results.  No method
exhibits a noticeable performance gain over the \ar~method, leading us to
dismiss the implicit regularization hypothesis.

@@

In this section, we provide extra details about the up/down/flat analysis
described in the main manuscript. Figure \ref{fig:upswing-summary} shows the
overall results, examining the average difference $\WIS(AR) - \WIS(F)$ in
period. Figure \ref{fig:hotspots-upswing-downswing} shows the same information
for the hotspot task. On average, during downswings and flat periods, the
indicator models have lower classification error and higher log likelihood than
the AR model. For hotspots, both Google-AA and CTIS-CLIIC perform better than
the AR model during upswings, in contrast to the forecasting task, where only
Google-AA improves. For a related analysis, Figure \ref{fig:cor-wis-ratio} shows
histograms of the Spearman correlation (Spearman's $\rho$, a rank-based measure
of association) between the $\WIS(F)/\WIS(AR)$ and the magnitude of the
swing. Again we see that case rate increases are positively related to
diminished performance of the indicator models.

One hypothesis for diminished relative performance during upswings is that the
AR model tends to overpredict downswings and underpredict upswings. Adding
indicators appears to help avoid this behavior on the downswing but not as much
on upswings. Figure \ref{fig:upswing-corr-table} shows the correlation between
between $\WIS(AR) - \WIS(F)$ and the difference of their median forecasts.
During downswings, this correlation is large, implying that improved relative
performance of $F$ is related to making lower forecasts than the AR model. The
opposite is true during upswings. This is largely to be expected. However, the
relationship attenuates in flat periods and during upswings. That is, when
performance is better in those cases, it may be due to other factors than simply
making predictions in the correct direction, for example, narrower confidence
intervals.

It is important to note, that even though some indicators, notably CHNG-COVID 
and CHNG-CLI underperform relative to the AR model during upswings, all models 
dramatically outperform the baseline in such periods. Figure 
\ref{fig:upswing-summary-remake} shows the performance of all forecasters relative
to the null model. All forecasters suffer relative to the baseline during down 
periods, but the AR is the worst. In contrast, all models beat the baseline
during up periods, even CHNG-COVID and CHNG-CLI, though not by quite as much as the
AR does.

@@

In Section 2.D of the main text, we discuss the extent to which the indicators
are leading or lagging case rates during different periods. To define the amount
of leadingness or laggingness at location $\ell$, we use the cross correlation
function (CCF) between the two time series. The $\CCF_\ell(a)$ of an indicator
$X_{\ell}$ and case rates $Y_{\ell}$ is defined as
% \[
% \CCF_{\ell,t}(a) = \frac{1}{n_a} \sum_{i=1}^{n_a} \left( ( X_{\ell,t,i+a} -
%   \overline{X}_{\ell,t}) / s^{X}_{\ell,t}\right) \left( ( Y_{\ell,t,i} -
%   \overline{Y}_{\ell,t}) / s^{Y}_{\ell,t}\right), 
% \]
% where $n_a$ is the number of available time points when $X_{\ell,t}$ has been
% shifted in time by $a$ days, and $\overline{X}_{\ell,t}$, $s^X_{\ell,t}$
% are the sample mean and
% standard deviation of $X_{\ell,t}$ (respectively $Y_{\ell,t}$). The result is a
their Pearson correlation where $X_\ell$ has been aligned with the values of
$Y_{\ell}$ that occurred $a$ days earlier.  Thus, for any $a>0$,
$\CCF_{\ell}(a)>0$ indicates that $Y_{\ell,t}$ is moving together with
$X_{\ell,t+a}$.  In this case we say that $X_{\ell}$ is lagging $Y_{\ell}$. For
$a<0$, $\CCF_{\ell}(a)>0$ means that $Y_{\ell,t}$ is positively correlated with
$X_{\ell,t-a}$, so we say that $X_{\ell}$ leads $Y_{\ell}$.

Figure \ref{fig:ccf-dv-finalized} shows the standardized signals for the HRR
containing Charlotte, North Carolina, from August 1, 2020 until the end of
September. These are the same signals shown in Figure 1 in the manuscript but
using finalized data. To define ``leadingness'' we compute $\CCF_{\ell}(a)$ (as
implemented with the R function {\tt ccf()}) for each $a\in \{-15,\ldots,15\}$
using the 56 days leading up to the target date. This is the same amount of data
used to train the forecasters: 21 days of training data, 21 days to get the
response at $a=21$, and 14 days for the longest lagged value. The orange dashed
horizontal line represents the 95\% significance threshold for correlations
based on 56 observations. Any correlations larger in magnitude than this value
are considered statistically significant under the null hypothesis of no
relationship. We define leadingness to be the sum of the significant
correlations that are leading (those above the dashed line with $a<0$) while
laggingness is the same but for $a>0$. In the figure, there are three
significant correlations on the ``leading'' side (at $a = -5, -4, -3$), so
leadingness will be the sum of those values while laggingness is 0: on September
28 in Charlotte, \dv~is leading cases leading but not lagging.

Figure \ref{fig:lagging-only} shows the correlation between laggingness and the
difference in indicator WIS and AR WIS. Unlike leadingness (Figure 5 in the
manuscript) there is no obvious relationship that holds consistently across
indicators. This is heartening as laggingness should not aid forecasting
performance. On the other hand, if an indicator is more lagging than it is
leading, this may suggest diminished performance.
Figure~\ref{fig:diff-in-lead-lag} shows the correlation of the difference in
leadingness and laggingness with the difference in WIS. The pattern here is
largely similar to the pattern in leadingness described in the manuscript: the
relationship is strongest in down periods and weakest in up periods with the
strength diminishing as we move from down to flat to up for all indicators.

In calculating the CCF and the associated leadingness and laggingness scores, we
have used the finalized data, and we look at the behavior at the target date of
the forecast. That is we are using the same data to evaluate predictive accuracy
as to determine leadingness and laggingness. It should be noted that the
leadingness of the indicator at the time the model is trained may also be
important. Thus, we could calculate separate leadingness and laggingness scores
for the trained model and for the evaluation data and examine their combination
in some way. We do not pursue this combination further and leave this
investigation for future work.

% \section{Examining data in 2021}

% In this section, we investigate the sensitivity of the results to the
% period over which we train and evaluate the models.  In the main
% paper, we end all evaluation on December 31, 2020.  Figures
% \ref{fig:fcast-alldates}--\ref{fig:hot-alldates} show how the
% results would differ if we extended this analysis through March
%   31, 2021. Comparing Figure \ref{fig:fcast-alldates} to Figure 3 of
% the main paper, one sees that as ahead increases most methods now improve
% relative to the baseline forecaster. When compared to other methods, 
% \chngcli~appears much better than
% it had previously; however, all forecasters other than \chngcov~and
% \dv~are performing less well relative to the baseline than before.
% These changes are likely due to the differing nature of the pandemic
% in 2021, with flat and downward trends much more common than upward
% trajectories.  Indeed, the nature of the hotspot prediction problem is
% quite different in this period.  With a 21-day training window, it is
% common for there to be many fewer hotspots in training.

\section{Aggregation Over Time and Space}

Following the suggestion of an anonymous reviewer, we investigate two
other disaggregated versions of the main forecasting result shown in Figure 3 on the
manuscript. The first (Figure \ref{fig:cumulative-mean}) displays the cumulative
error (summed over all horizons) of each forecaster divided by the cumulative
error of the baseline.  This perspective should illustrate how the models
perform over time, drawing attention to any nonstationary behavior.  During the
initial increase in cases in July 2020, CTIS-CLIIC and Google-AA gain a lot
relative to the AR model. All the indicators do much better than the AR model
during the following downturn (the ebb of the second wave). The AR model
actually improves over the indicators in October 2020, before losing a bit in
late November. The second (Figure \ref{fig:cumulative-geo-mean}) repeats this 
analysis but aggregating by Geometric mean as described above and displays a
similar pattern.

Figure \ref{fig:errs-in-space} examines the spatial behavior over the entire
period of the indicator models relative to the AR. For ease of
comparison, we show the percent improvement in each HRR. Negative numbers (blue)
mean that the indicator helped while positives (red) mean that the indicator
hurt forecasting performance. The clear pattern is that in most HRRs, the
indicators improved performance, though usually by small amounts (2.5\%--10\%).
In some isolated HRRs, performance was markedly worse, though there does not
appear to be any particular pattern to these locations.

>>>>

As explained in Section 2.B of the main paper, a (somewhat cynical) hypothesis
for why we see benefits in forecasting and hotspot prediction is that the
indicators are not actually providing useful information but they are instead
acting as a sort of ``implicit regularization,'' leading to shrinkage on the
autoregressive coefficients and therefore to less volatile predictions.  To
investigate this hypothesis, we consider fitting ``noise features'' that in
truth have zero relationship to the response.  Recall (from the main paper) that
at each forecast date, we train a model on 6,426 location-time pairs.  Each
indicator model uses 6 features, corresponding to the 3 autoregressive terms and
the 3 lagged indicator values.  To form noise indicator features, we replace
their values with those from a randomly chosen time-space pair (while keeping
the autoregressive features fixed).  In particular, at each location $\ell$ and
time $t$, for the forecasting task we replace the triplet $(X_{\ell,t},
X_{\ell,t-7}, X_{\ell,t-14})$ in Eq.\ (3) of the main
paper with the triplet \smash{$(X_{\ell^*,t^*}, X_{\ell^*,t^*-7},
  X_{\ell^*,t^*-14})$}, where $(\ell^*,t^*)$ is a location-time pair sampled with
replacement from the 6,426 total location-time pairs.  Likewise in the hotspot 
prediction task, we replace the triplet \smash{$(X_{\ell,t}^\Delta,
  X_{\ell,t-7}^\Delta, X_{\ell,t-14}^\Delta)$} in Eq.\ (5) of the main paper with
\smash{$(X_{\ell^*,t^*}^\Delta, X_{\ell^*,t^*-7}^\Delta,
  X_{\ell^*,t^*-14}^\Delta)$}.  Figures
\ref{fig:fcast-booted}--\ref{fig:hot-booted} show the results.  No method
exhibits a noticeable performance gain over the \ar~method, leading us to
dismiss the implicit regularization hypothesis.

@@

In this section, we provide extra details about the upswing/downswing analysis
described in the main manuscript, Section 2.C. Figure \ref{fig:upswing-summary}
shows the overall results, examining the average difference $\WIS(\mathrm{AR}) - 
\WIS(F)$ for each forecaster $F$, in each in each period.  Figure
\ref{fig:hotspots-upswing-downswing} shows the same information for the hotspot 
task.  On average, during downswings and flat periods, the indicator models have
lower classification error and higher log likelihood than the AR model.  For
hotspots, both Google-AA and CTIS-CLIIC perform better than the AR model during
upswings, in contrast to the forecasting task, where only Google-AA improves.
In a related analysis, Figure \ref{fig:cor-wis-ratio} shows histograms of the
Spearman correlation (Spearman's $\rho$, a rank-based measure of association)
between the ratio $\WIS(F)/\WIS(\mathrm{AR})$ and the magnitude of the swing.
Again we see that case rate increases are positively related to diminished
performance of the indicator models.

One hypothesis for diminished relative performance during upswings is that the
AR model tends to overpredict downswings and underpredict upswings.  Adding
indicators appears to help avoid this behavior on the downswing but not as much
on upswings.  Figure \ref{fig:upswing-corr-table} shows the correlation between
between $\WIS(\mathrm{AR}) - \WIS(F)$ and the difference of their median
forecasts.  During downswings, this correlation is large, implying that improved
relative performance of $F$ is related to making lower forecasts than the AR
model.  The opposite is true during upswings.  This is largely to be expected.
However, the relationship attenuates in flat periods and during upswings. That
is, when performance is better in those cases, it may be due to other factors
than simply making predictions in the correct direction, for example, narrower
confidence intervals.

It is important to note that, even though some indicators---notably CHNG-COVID
and CHNG-CLI---underperform relative to the AR model during upswings, all models
dramatically outperform the baseline in such periods.  Furthermore, Figure
\ref{fig:upswing-summary-remake} shows the performance of all forecasters
relative to the baseline model.  All forecasters suffer relative to the baseline
during down periods, but the AR is the worst. In contrast, all models beat the
baseline during up periods, even CHNG-COVID and CHNG-CLI, though not by quite as
much as the AR does.

@@

In Section 2.D of the main text, we discuss the extent to which the indicators
are leading or lagging case rates during different periods. To define the amount
of leadingness or laggingness at location $\ell$, we use the cross correlation
function (CCF) between the two time series. The $\CCF_\ell(a)$ of an indicator
series $X_{\ell}$ and case rate series $Y_{\ell}$ is defined as their Pearson 
correlation where $X_\ell$ has been aligned with the values of $Y_{\ell}$ that
occurred $a$ days earlier.  Thus, for any $a>0$, $\CCF_{\ell}(a)>0$ indicates
that $Y_{\ell,t}$ is moving together with $X_{\ell,t+a}$.  In this case we say
that $X_{\ell}$ is lagging $Y_{\ell}$. For $a<0$, $\CCF_{\ell}(a)>0$ means that
$Y_{\ell,t}$ is positively correlated with $X_{\ell,t-a}$, so we say that
$X_{\ell}$ leads $Y_{\ell}$.

Figure \ref{fig:ccf-dv-finalized} shows the standardized signals for the HRR
containing Charlotte, North Carolina, from August 1, 2020 until the end of
September. These are the same signals shown in Figure 1 in the manuscript, but 
using finalized data. To define ``leadingness'' we compute $\CCF_{\ell}(a)$ (as
implemented with the R function {\tt ccf()}) for each $a\in \{-15,\ldots,15\}$
using the 56 days leading up to the target date. This is the same amount of data
used to train the forecasters: 21 days of training data, 21 days to get the
response at $a=21$, and 14 days for the longest lagged value. The orange dashed
horizontal line represents the 95\% significance threshold for correlations
based on 56 observations. Any correlations larger in magnitude than this value
are considered statistically significant under the null hypothesis of no
relationship. We define leadingness to be the sum of the significant
correlations that are leading (those above the dashed line with $a<0$) while
laggingness is the same but for $a>0$. In the figure, there are three
significant correlations on the ``leading'' side (at $a = -5, -4, -3$), so
leadingness will be the sum of those values while laggingness is 0: on September
28 in Charlotte, \dv~is leading cases leading but not lagging.

Figure \ref{fig:lagging-only} shows the correlation between laggingness and the
difference in forecaster WIS and AR WIS. Unlike leadingness (Figure 5 in the
manuscript) there is no obvious relationship that holds consistently across
indicators. This is encouraging as laggingness should not aid forecasting 
performance. On the other hand, if an indicator is more lagging than it is
leading, this may suggest diminished performance.
Figure \ref{fig:diff-in-lead-lag} shows the correlation of the difference in
leadingness and laggingness with the difference in WIS. The pattern here is
largely similar to the pattern in leadingness described in the manuscript: the
relationship is strongest in down periods and weakest in up periods with the
strength diminishing as we move from down to flat to up for all indicators.

In calculating the CCF and the associated leadingness and laggingness scores, we
have used the finalized data, and we look at the behavior at the target date of
the forecast. That is we are using the same data to evaluate predictive accuracy
as to determine leadingness and laggingness. It should be noted that the
leadingness of the indicator at the time the model is trained may also be
important.  Thus, we could calculate separate leadingness and laggingness scores
for the trained model and for the evaluation data and examine their combination
in some way.  However, we do not pursue this further.

\section{Disaggregation Over Time and Space}

Following the suggestion of an anonymous reviewer, we investigate two
other disaggregated versions of the main forecasting result shown in Figure 3 on the
manuscript. Below we use the term ``error'' to refer to the WIS summed over all
ahead values.  The first (Figure \ref{fig:cumulative-mean}) displays the
cumulative error, up through any point in time, of each forecaster divided by
the cumulative error of the baseline.  This perspective should illustrate how
the models perform over time, drawing attention to any nonstationary behavior.
During the initial increase in cases in July 2020, CTIS-CLIIC and Google-AA gain
quite a bit acurracy compared to the AR model.  All the indicators do much
better than the AR model during the following downturn (the ebb of the second 
wave). The AR model actually improves over the indicators in October 2020,
before losing a bit in late November. % The second (Figure
% \ref{fig:cumulative-geo-mean}) repeats this analysis but aggregating by 
% geometric mean, and presents a similar message.

Figure \ref{fig:errs-in-space} examines the spatial behavior of errors over the
entire period of the indicator models relative to the AR.  For ease of
comparison, we show the percent improvement in each HRR.  Negative numbers
(blue) mean that the indicator helped while positives (red) mean that the
indicator hurt forecasting performance.  The clear pattern is that in most HRRs,
the indicators improved performance, though usually by small amounts
(2.5\%--10\%).  In some isolated HRRs, performance was markedly worse, though
there does not appear to be any particular pattern to these locations.
Interestingly, the geographic patterns of improvement differ quite a lot in
between the indicators.  This suggests that a forecasting model that carefully
combines all of indicators could be a considerable improvement.
